---
author:  Termeh Shafie
title: "Social Network Analysis"
subtitle: "Worksheet 2: Graph Theory"
lightbox: auto
toc: true
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 10
---

## Graph Theory in R with `igraph`

We'll use the `igraph` package to explore key graph-theoretical concepts.

```{r setup}
#| message: FALSE
library(igraph)
```

## Undirected Graph

We'll use an undirected graph with 8 nodes.

```{r}
g <- graph_from_literal(
  1 -- 2, 
  1 -- 3,
  2 -- 4,
  3 -- 5,
  4 -- 5,
  5 -- 6,
  6 -- 7,
  6 -- 8,
  7 -- 8
)

plot(g, vertex.label.cex = 1.2, vertex.size = 20)
```

------------------------------------------------------------------------

###  Degree & Degree Distribution

```{r}
deg <- degree(g)
deg

# Degree distribution
dist <- degree_distribution(g)
plot(dist, type = "h", main = "Degree Distribution", xlab = "Degree", ylab = "Frequency")
```

Identify nodes with the highest and lowest degree. How does the distribution look?

------------------------------------------------------------------------


### Graph Diameter

```{r}
diameter(g)
```

What is the longest shortest path in the graph?

------------------------------------------------------------------------

### Shortest Paths

Find shortest paths from node 1 to all others:

```{r}
sp <- distances(g, v = 1)
sp
```

 Do you understand the output? What is the shortest path from node 1 to node 6?

------------------------------------------------------------------------


### Adjacency Matrix

```{r}
adj_matrix <- as_adjacency_matrix(g, sparse = FALSE)
adj_matrix
```

Use the adjacency matrix to compute the degree of each node. Compare with `degree(g)`.

------------------------------------------------------------------------


### Cutpoints (Articulation Points)

```{r}
articulation_points <- articulation_points(g)
articulation_points
```

### Bridges (Critical Edges)

```{r}
bridge_edges <- which(is.na(edge_connectivity(g)))
E(g)[bridge_edges]

# Alternatively use:
bridges(g)
```

### Visualize Cutpoints and Bridges

```{r}
V(g)$color <- ifelse(V(g) %in% articulation_points, "red", "skyblue")
E(g)$color <- ifelse(E(g) %in% bridges(g), "red", "black")

plot(g, vertex.size = 20, vertex.label.cex = 1.2)
```

 Which nodes and edges are critical to keeping the graph connected?

------------------------------------------------------------------------

## Directed Graph

Now let’s work with a **directed graph**.

```{r}
g_dir <- graph_from_literal(
  A -+ B, A -+ C,
  B -+ D,
  C -+ D,
  D -+ E,
  E -+ F,
  F -+ C
)

plot(g_dir, vertex.label.cex = 1.2, vertex.color = "lightcoral", edge.arrow.size = 0.5)
```


### In-Degree and Out-Degree

```{r}
degree(g_dir, mode = "in")   # incoming links
degree(g_dir, mode = "out")  # outgoing links
```

### Strongly Connected Components

```{r}
components(g_dir, mode = "strong")
```

### Directed Paths and Diameter

```{r}
diameter(g_dir, directed = TRUE)
```

Explore how cycles and direction affect path lengths.

------------------------------------------------------------------------


## Exercises

- Create some other small unidrected and directed graphs and see how the above measure vary on them

- import the Florentine marriage and business network from the `networkdata` package and compute the appropriate measures from above on it
```{r}
library(networkdata)
data("flo_marriage")
data("flo_business")
```
