---
author:  Termeh Shafie
title: "Social Network Analysis"
subtitle: "Worksheet 11: SAOMs I"
lightbox: auto
toc: true
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 10
execute: 
  cache: true
  warning: false
  message: false
---

# Introduction

In this practical we will estimate a **Stochastic Actor-Oriented Model (SAOM)** using the **Teenage Friends and Lifestyle Study**.

The data consist of:

- Three waves of friendship nominations among 129 pupils.
- Individual characteristics such as sex.
- Smoking behavior measured at each wave.

The goal is to learn how friendship networks evolve over time and how smoking behavior co-evolves with friendships.


## Load packages

```{r}
library(igraph)
library(networkdata)
library(RSiena)
```


## Load the data

The dataset contains three friendship networks measured over time.

```{r}
data("glasgow129")

glasgow_g1 <- glasgow129[[1]]
glasgow_g2 <- glasgow129[[2]]
glasgow_g3 <- glasgow129[[3]]
```


## Prepare the friendship networks

RSiena requires the network to be stored as a three-dimensional array:

- rows = sender
- columns = receiver
- third dimension = observation wave

```{r}
net1 <- as_adjacency_matrix(glasgow_g1, sparse = FALSE)
net2 <- as_adjacency_matrix(glasgow_g2, sparse = FALSE)
net3 <- as_adjacency_matrix(glasgow_g3, sparse = FALSE)

friendshipArray <- array(
  c(net1, net2, net3),
  dim = c(129,129,3)
)
```



# Model 1: Network dynamics only

We begin with a simple model that explains changes in friendship **using only structural network effects**.


# Creating the dependent network

The friendship network is the dependent variable.

The function `sienaDependent()` tells RSiena that this network changes over time and should be modelled.

```{r}
friendship <- sienaDependent(friendshipArray)
```

First we create a Siena data object.

```{r}
mydata1 <- sienaDataCreate(friendship)
```

Next we create an object that will contain the effects to estimate.

```{r}
myeff1 <- getEffects(mydata1)
```

Let's inspect the default effects.

```{r}
myeff1
```


## Adding structural effects

We now add three common network mechanisms.

```{r}
myeff1 <- includeEffects(
  myeff1,
  outdegree,
  recip,
  transTrip
)
```

These effects represent:

- **Outdegree (density):** baseline tendency to create friendships.
- **Reciprocity:** tendency to reciprocate friendships.
- **Transitive triplets:** tendency for friends of friends to become friends.


## Estimate the model

The estimation algorithm first needs to be specified.

```{r}
myalgorithm1 <- sienaAlgorithmCreate(
  projname="SAOM1",
  seed=12345
)
```

Now estimate the model.

```{r}
result1 <- siena07(
  myalgorithm1,
  data=mydata1,
  effects=myeff1,
  batch=TRUE
)
```

Display the results.

```{r}
summary(result1)
```


# Model 2: Adding actor covariates

Next we investigate whether friendships depend on:

- sex
- smoking

## Sex

The sex variable does not change over time.

Therefore it is entered as a **constant covariate**.

```{r}
sex <- ifelse(
  V(glasgow_g2)$sex.F=="F",
  1,
  0
)

sexCov <- coCovar(sex)
```

## Smoking

Smoking is observed at every wave.

We therefore create a changing covariate.

```{r}
smoke <- cbind(
  as.numeric(V(glasgow_g1)$tobacco),
  as.numeric(V(glasgow_g2)$tobacco),
  as.numeric(V(glasgow_g3)$tobacco)
)

smokeCov <- varCovar(smoke)
```



## Create the new Siena data object

```{r}
mydata2 <- sienaDataCreate(
  friendship,
  sexCov,
  smokeCov
)
```

```{r}
myeff2 <- getEffects(mydata2)
```


## Add covariate effects

For each covariate we estimate three effects.

```{r}
myeff2 <- includeEffects(
  myeff2,
  egoX,
  altX,
  simX,
  interaction1="sexCov"
)

myeff2 <- includeEffects(
  myeff2,
  egoX,
  altX,
  simX,
  interaction1="smokeCov"
)
```

These correspond to:

- **ego effect:** Does this type of actor nominate more friends?
- **alter effect:** Are actors with this characteristic more popular?
- **similarity effect:** Do similar actors become friends?



## Estimate Model 2

```{r}
myalgorithm2 <- sienaAlgorithmCreate(
  projname="SAOM2",
  seed=12345
)

result2 <- siena07(
  myalgorithm2,
  data=mydata2,
  effects=myeff2,
  batch=TRUE
)

summary(result2)
```


# Model 3: Co-evolution of friendship and smoking

So far smoking has only been treated as a covariate.

Now we allow smoking itself to change over time.

We therefore model **two dependent variables simultaneously**:

- friendship
- smoking behavior

```{r}
smoking <- sienaDependent(
  smoke,
  type="behavior"
)
```


## Create the co-evolution dataset

```{r}
mydata3 <- sienaDataCreate(
  friendship,
  smoking,
  sexCov
)

myeff3 <- getEffects(mydata3)
```



## Friendship dynamics

We include the same structural effects as before.

```{r}
myeff3 <- includeEffects(
  myeff3,
  outdegree,
  recip,
  transTrip,
  name="friendship"
)
```


## Behavior dynamics

We now model changes in smoking.

```{r}
myeff3 <- includeEffects(
  myeff3,
  linear,
  quad,
  name="smoking"
)
```

The behavioral objective function now estimates:

- **Linear shape:** Is higher smoking generally more or less attractive?
- **Quadratic shape:** Are moderate or extreme smoking levels preferred?



## Social influence

Finally we estimate peer influence.

```{r}
myeff3 <- includeEffects(
  myeff3,
  avAlt,
  name="smoking",
  interaction1="friendship"
)
```

This tests whether students tend to become more similar to the average smoking behavior of their friends.



## Estimate the co-evolution model

```{r}
myalgorithm3 <- sienaAlgorithmCreate(
  projname="SAOM3",
  seed=12
)

result3 <- siena07(
  myalgorithm3,
  data = mydata3,
  effects = myeff3,
  batch = TRUE,
  returnDeps = TRUE
)

summary(result3)
```

::: {.callout-important}
To use `sienaGOF()`which we do in the next seection, the model must be estimated with `returnDeps = TRUE`, so that RSiena stores the simulated dependent variables needed for the goodness-of-fit checks.
:::


   
# Optional: nicer tables

```{r}
#| eval: false 

siena_table(result1, type = "html", sig = TRUE)
siena_table(result2, type = "html", sig = TRUE)
siena_table(result3, type = "html", sig = TRUE)
```



## Goodness of Fit

Estimating a SAOM is only the first step. As with Exponential Random Graph Models (ERGMs), it is good practice to assess **how well the fitted model reproduces the observed data**.

The basic idea is to:

1. Simulate many networks from the fitted model.
2. Compare characteristics of the simulated networks with those of the observed network.
3. Evaluate whether the observed network looks typical under the fitted model.

If the observed network falls well within the distribution of simulated networks, this provides evidence that the model adequately captures that aspect of the data.

Formally, the goodness-of-fit assessment tests the **null hypothesis that the fitted SAOM is capable of generating networks with the same structural properties as the observed network**. The plots compare the observed statistic (typically shown as a vertical line or point) with the distribution of the same statistic across networks simulated from the fitted model. The accompanying $p$-value quantifies how unusual the observed statistic is under the fitted model. A large $p$-value (commonly greater than 0.05) indicates that the observed network is consistent with the simulations, suggesting an adequate fit for that network feature. A small $p$-value indicates that the observed statistic lies in the tails of the simulated distribution, implying that the model does not reproduce that particular aspect of the network well and may require additional effects or refinements.



### Goodness of fit: Indegree distribution

First, we assess whether the fitted model reproduces the **indegree distribution** (i.e., how many friendship nominations each student receives).

The function `sienaGOF()` requires:

- `sienaFitObject`: the fitted SAOM.
- `auxiliaryFunction`: the network statistic to assess.
- `varName`: the dependent network or behavior.

```{r}
#| message: false

gof_indegree <- sienaGOF(
  result3,
  IndegreeDistribution,
  varName = "friendship"
)

plot(gof_indegree)
```

Interpret the figure by comparing the observed indegree distribution (solid line) with the distribution obtained from the simulated networks (boxplots).


### Goodness of fit: Smoking behavior

Next, we assess whether the model reproduces the observed distribution of smoking behavior.

```{r}
#| message: false

gof_behavior <- sienaGOF(
  result3,
  BehaviorDistribution,
  varName = "smoking"
)

plot(gof_behavior)
```

Again, compare the observed behavior distribution with the simulated distributions.

::: {.callout-note}
## Why assess goodness of fit for smoking behaviour?

Although smoking behaviour is modelled directly in Model 3, it is still useful to assess its goodness of fit. The estimated parameters are chosen to explain **how smoking changes over time**, not to perfectly reproduce every feature of the observed data.

A behavioural goodness-of-fit check asks whether simulations from the fitted model generate a distribution of smoking behaviour similar to the one observed in the data. If the simulated and observed distributions are similar, this provides evidence that the model adequately captures the overall behavioural dynamics. If they differ substantially, the model may be missing important behavioural processes, even if individual parameter estimates appear reasonable.

The parameters in a SAOM define an actor-oriented objective function governing the micro-steps of network and behavioural change. The estimation procedure is not matching the final distribution of smoking behaviour in the same way an ERGM matches sufficient statistics. In other words, the ERGM is matching network statistics directly, whereas the SAOM is estimating a dynamic process whose consequences can still be checked using GOF diagnostics.
:::



## Discussion Questions

After estimating **Model 3 (Co-evolution of Friendship and Smoking)**, answer the following questions.

1. **Friendship dynamics:** Is there evidence that friendship ties are shaped by **reciprocity** and **transitive triplets**? Interpret the estimated coefficients for these structural effects.

2. **Network structure:** What does the **negative outdegree (density)** parameter imply about the overall tendency to form friendships in the network?

3. **Behavior dynamics:** What do the **linear shape** and **quadratic shape** parameters indicate about changes in smoking behavior over time? Do they suggest that students tend to move toward lower, higher, or moderate levels of smoking?

4. **Social influence:** Does the **average alter (`avAlt`)** effect provide evidence that students' smoking behavior is influenced by the smoking behavior of their friends? Interpret both the sign and magnitude of this parameter.

5. **Statistical significance:** Which estimated parameters appear to be statistically significant (using the estimates and standard errors)? What do these significant effects tell us about friendship evolution and smoking behavior?

6. **Model convergence:** Assess the convergence of the model.

   - What is the overall maximum convergence ratio?

   - Is it below the recommended threshold (preferably < 0.10, acceptable < 0.25)?

   - Based on the convergence diagnostics, would you consider the parameter estimates reliable? Explain your reasoning.

   - If convergence is poor, what changes to the model or estimation procedure might improve it?

7. **Goodness of fit (Network):** Does the fitted model adequately reproduce the **indegree distribution**? Use the goodness-of-fit plot to justify your answer.

8. **Goodness of fit (Behavior):** Does the fitted model adequately reproduce the **distribution of smoking behavior**? What aspects of the observed distribution are well captured, and where does the model perform poorly?

9. **Overall assessment:** Considering the parameter estimates, convergence diagnostics, and goodness-of-fit results, how well does the co-evolution SAOM explain the joint evolution of friendship ties and smoking behavior? What limitations of the model remain?