---
author:  Termeh Shafie
title: "Social Network Analysis"
subtitle: "Worksheet 11: SAOMs I"
lightbox: auto
toc: true
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 10
execute: 
  cache: true
  warning: false
  message: false
---

# Introduction

In this practical we will estimate a **Stochastic Actor-Oriented Model (SAOM)** using the **Teenage Friends and Lifestyle Study**.

The data consist of:

- Three waves of friendship nominations among 129 pupils.
- Individual characteristics such as sex.
- Smoking behavior measured at each wave.

The goal is to learn how friendship networks evolve over time and how smoking behavior co-evolves with friendships.


## Load packages

```{r}
library(igraph)
library(networkdata)
library(RSiena)
```


## Load the data

The dataset contains three friendship networks measured over time.

```{r}
data("glasgow129")

glasgow_g1 <- glasgow129[[1]]
glasgow_g2 <- glasgow129[[2]]
glasgow_g3 <- glasgow129[[3]]
```


## Prepare the friendship networks

RSiena requires the network to be stored as a three-dimensional array:

- rows = sender
- columns = receiver
- third dimension = observation wave

```{r}
net1 <- as_adjacency_matrix(glasgow_g1, sparse = FALSE)
net2 <- as_adjacency_matrix(glasgow_g2, sparse = FALSE)
net3 <- as_adjacency_matrix(glasgow_g3, sparse = FALSE)

friendshipArray <- array(
  c(net1, net2, net3),
  dim = c(129,129,3)
)
```



# Model 1: Network dynamics only

We begin with a simple model that explains changes in friendship **using only structural network effects**.


# Creating the dependent network

The friendship network is the dependent variable.

The function `sienaDependent()` tells RSiena that this network changes over time and should be modelled.

```{r}
friendship <- sienaDependent(friendshipArray)
```

First we create a Siena data object.

```{r}
mydata1 <- sienaDataCreate(friendship)
```

Next we create an object that will contain the effects to estimate.

```{r}
myeff1 <- getEffects(mydata1)
```

Let's inspect the default effects.

```{r}
myeff1
```


## Adding structural effects

We now add three common network mechanisms.

```{r}
myeff1 <- includeEffects(
  myeff1,
  outdegree,
  recip,
  transTrip
)
```

These effects represent:

- **Outdegree (density):** baseline tendency to create friendships.
- **Reciprocity:** tendency to reciprocate friendships.
- **Transitive triplets:** tendency for friends of friends to become friends.


## Estimate the model

The estimation algorithm first needs to be specified.

```{r}
myalgorithm1 <- sienaAlgorithmCreate(
  projname="SAOM1",
  seed=12345
)
```

Now estimate the model.

```{r}
result1 <- siena07(
  myalgorithm1,
  data=mydata1,
  effects=myeff1,
  batch=TRUE
)
```

Display the results.

```{r}
summary(result1)
```


# Model 2: Adding actor covariates

Next we investigate whether friendships depend on:

- sex
- smoking

## Sex

The sex variable does not change over time.

Therefore it is entered as a **constant covariate**.

```{r}
sex <- ifelse(
  V(glasgow_g2)$sex.F=="F",
  1,
  0
)

sexCov <- coCovar(sex)
```

## Smoking

Smoking is observed at every wave.

We therefore create a changing covariate.

```{r}
smoke <- cbind(
  as.numeric(V(glasgow_g1)$tobacco),
  as.numeric(V(glasgow_g2)$tobacco),
  as.numeric(V(glasgow_g3)$tobacco)
)

smokeCov <- varCovar(smoke)
```



## Create the new Siena data object

```{r}
mydata2 <- sienaDataCreate(
  friendship,
  sexCov,
  smokeCov
)
```

```{r}
myeff2 <- getEffects(mydata2)
```


## Add covariate effects

For each covariate we estimate three effects.

```{r}
myeff2 <- includeEffects(
  myeff2,
  egoX,
  altX,
  simX,
  interaction1="sexCov"
)

myeff2 <- includeEffects(
  myeff2,
  egoX,
  altX,
  simX,
  interaction1="smokeCov"
)
```

These correspond to:

- **ego effect:** Does this type of actor nominate more friends?
- **alter effect:** Are actors with this characteristic more popular?
- **similarity effect:** Do similar actors become friends?



## Estimate Model 2

```{r}
myalgorithm2 <- sienaAlgorithmCreate(
  projname="SAOM2",
  seed=12345
)

result2 <- siena07(
  myalgorithm2,
  data=mydata2,
  effects=myeff2,
  batch=TRUE
)

summary(result2)
```


# Model 3: Co-evolution of friendship and smoking

So far smoking has only been treated as a covariate.

Now we allow smoking itself to change over time.

We therefore model **two dependent variables simultaneously**:

- friendship
- smoking behavior

```{r}
smoking <- sienaDependent(
  smoke,
  type="behavior"
)
```


## Create the co-evolution dataset

```{r}
mydata3 <- sienaDataCreate(
  friendship,
  smoking,
  sexCov
)

myeff3 <- getEffects(mydata3)
```



## Friendship dynamics

We include the same structural effects as before.

```{r}
myeff3 <- includeEffects(
  myeff3,
  outdegree,
  recip,
  transTrip,
  name="friendship"
)
```


## Behavior dynamics

We now model changes in smoking.

```{r}
myeff3 <- includeEffects(
  myeff3,
  linear,
  quad,
  name="smoking"
)
```

The behavioral objective function now estimates:

- **Linear shape:** Is higher smoking generally more or less attractive?
- **Quadratic shape:** Are moderate or extreme smoking levels preferred?



## Social influence

Finally we estimate peer influence.

```{r}
myeff3 <- includeEffects(
  myeff3,
  avAlt,
  name="smoking",
  interaction1="friendship"
)
```

This tests whether students tend to become more similar to the average smoking behavior of their friends.



## Estimate the co-evolution model

```{r}
myalgorithm3 <- sienaAlgorithmCreate(
  projname="SAOM3",
  seed=12
)

result3 <- siena07(
  myalgorithm3,
  data = mydata3,
  effects = myeff3,
  batch = TRUE,
  returnDeps = TRUE
)

summary(result3)
```

::: {.callout-important}
To use `sienaGOF()`which we do in the next seection, the model must be estimated with `returnDeps = TRUE`, so that RSiena stores the simulated dependent variables needed for the goodness-of-fit checks.
:::


   
# Optional: nicer tables

```{r}
#| eval: false 

siena_table(result1, type = "html", sig = TRUE)
siena_table(result2, type = "html", sig = TRUE)
siena_table(result3, type = "html", sig = TRUE)
```



## Goodness of Fit

Estimating a SAOM is only the first step. As with Exponential Random Graph Models (ERGMs), it is good practice to assess **how well the fitted model reproduces the observed data**.

The basic idea is to:

1. Simulate many networks from the fitted model.
2. Compare characteristics of the simulated networks with those of the observed network.
3. Evaluate whether the observed network looks typical under the fitted model.

If the observed network falls well within the distribution of simulated networks, this provides evidence that the model adequately captures that aspect of the data.

Formally, the goodness-of-fit assessment tests the **null hypothesis that the fitted SAOM is capable of generating networks with the same structural properties as the observed network**. The plots compare the observed statistic (typically shown as a vertical line or point) with the distribution of the same statistic across networks simulated from the fitted model. The accompanying $p$-value quantifies how unusual the observed statistic is under the fitted model. A large $p$-value (commonly greater than 0.05) indicates that the observed network is consistent with the simulations, suggesting an adequate fit for that network feature. A small $p$-value indicates that the observed statistic lies in the tails of the simulated distribution, implying that the model does not reproduce that particular aspect of the network well and may require additional effects or refinements.



### Goodness of fit: Indegree distribution

First, we assess whether the fitted model reproduces the **indegree distribution** (i.e., how many friendship nominations each student receives).

The function `sienaGOF()` requires:

- `sienaFitObject`: the fitted SAOM.
- `auxiliaryFunction`: the network statistic to assess.
- `varName`: the dependent network or behavior.

```{r}
#| message: false

gof_indegree <- sienaGOF(
  result3,
  IndegreeDistribution,
  varName = "friendship"
)

plot(gof_indegree)
```

Interpret the figure by comparing the observed indegree distribution (solid line) with the distribution obtained from the simulated networks (boxplots).


### Goodness of fit: Smoking behavior

Next, we assess whether the model reproduces the observed distribution of smoking behavior.

```{r}
#| message: false

gof_behavior <- sienaGOF(
  result3,
  BehaviorDistribution,
  varName = "smoking"
)

plot(gof_behavior)
```

Again, compare the observed behavior distribution with the simulated distributions.


## Discussion Questions

After estimating the third model, answer the following questions:

1. Is there evidence for **reciprocity** in friendship nominations? Explain your answer using the estimated parameter.

2. Is friendship formation influenced by **similarity in smoking behavior**? If so, how would you interpret this effect?

3. Is **smoking behavior influenced by friends**? What does the estimated social influence effect suggest?

4. How would you interpret the **linear** and **quadratic** behavior parameters? What do they imply about students' preferences for different levels of smoking?

5. Which effects are **statistically significant**? Interpret their estimated coefficients in the context of friendship or smoking behavior.

6. Assess the **convergence** of the model.
   - What is the **overall maximum convergence ratio**?
   - Is it below the recommended threshold (e.g., **0.25**, preferably **0.10**)?
   - Based on the convergence ratios, would you consider the parameter estimates to be reliable? Explain your reasoning.
   
::: {.callout-tip}
**Hint:** If the model has not converged, consider possible reasons why. For example:

   - Would running more estimation iterations improve convergence?
   - Is the model too complex (e.g., too many effects)?
   - Could some effects be removed or added one at a time to identify problematic specifications?
   - Are there any unusually large parameter estimates that might indicate estimation problems?
:::
   

7. Does the fitted model adequately reproduce the **indegree distribution**? Explain your answer using the goodness-of-fit plot.

8. Does the fitted model adequately reproduce the **distribution of smoking behavior**?

9. Overall, what do the goodness-of-fit diagnostics suggest about the adequacy of the fitted SAOM? Are there any aspects of the data that the model appears to reproduce poorly?