---
author:  Termeh Shafie
title: "Social Network Analysis"
subtitle: "Worksheet 5b: Beyond 'Standard' Networks"
lightbox: auto
toc: true
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 10
---


# Signed Network and Two-Mode Networks

### Exercise 1: Correlates of War
Load the "Correlates of War" dataset from the `signnet` package. The “cowList” dataset contains a list of 52 signed networks of inter-state relations over time (1946-1999). Two countries are connected by a positive tie if they form an alliance or have a peace treaty. A negative tie exists between countries who are at war or in other kinds of conflicts. 

```{r load_signnet}
library(signnet)
data("cowList")
```

The dataset includes 51 networks of international relations between nations (aggregated on 3 year time windows). A positive tie indicates some form of alliance between countries and a negative tie conflict or war. 

Except for the first and last task, you can either do the exercises for all networks or choose one specific time window

- Did the number of conflicts increase or decrease over time?
- Before computing any balance scores: What would you expect in terms of structural balance?
- Calculate the triangle based balance score. Does it mach your intuition?
- Visualize a network using `ggsigned()` and decide whether it makes sense to estimate a regular blockmodel or if you need to specify a general one and compute the blockmodel


### Exercise 2: Two Mode Projections

The file `senate15.csv` contains all bill cosponsorships of the US Senate from 2015 to 2017. 

- read the file `dat <- read.csv("senate15.csv")`
- Create a data frame of the senators  
  `senators <- unique(dat[,1:2])`
- Construct the two mode network of senators and bills with  
  `g <- bipartite_from_data_frame(dat,"bill","name")`
- print the igraph object and check how the two modes are distinguished
- check `?bipartite.projection` and **only** create the weighted projection between senators. Call the network `proj`. What does the weight indicate?
- Inspect the plot. Can you see any structure?
- delete edges with a weight less than x, where you should try different values for x. What do you think are "good" values for x? Does a structure emerge for any x? 


:::{.callout-tip}
For the exploratory part of the exercise you can do the following:
```{r}
#| eval: false
V(proj)$party <- senators$party[match(V(proj)$name,senators$name)]
V(proj)$color <- ifelse(V(proj)$party=="D","blue",
                 ifelse(V(proj)$party=="R","red","yellow"))

plot(proj,layout = layout_with_kk,vertex.label = NA,vertex.size = 3)
```

to delete edges, use `delete_edges()` 

:::



