---
author:  Termeh Shafie
title: "Social Network Analysis"
subtitle: "Worksheet 2: Exploring Functions and Graph Theory"
lightbox: auto
toc: true
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 10
---

# Exploring packages and functions
*(Interactive Session)*

## Handling network data

### Inbuilt network data

The `networkdata` package includes around 1000 datsets and more than 2000 networks.
Throughout the course will use several examples using data from this package. Spend some time exploring datasets in this package (you will be asked to choose and work on one of them for you empirical study).

```{r show_data}
#| code-fold: false
data(package = "networkdata")
```

- Freeman’s datasets from <http://moreno.ss.uci.edu/data> (not available anymore)
- Movie networks from <https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/T4HBA3>
- Covert networks from <https://sites.google.com/site/ucinetsoftware/datasets/covert-networks>
- Animal networks from <https://bansallab.github.io/asnr/>
- Shakespeare’s plays networks build with data from <https://github.com/mallaham/Shakespeare-Plays>
- Some networks from <http://konect.uni-koblenz.de/>
- Tennis networks compiled from <https://github.com/JeffSackmann> (please give credit to him if you use this data)
- Star Wars Character Interactions (Episode 1-7) from <https://github.com/evelinag/StarWars-social-network>

## Graph Theory in R with `igraph`

We'll use the `igraph` package to explore key graph-theoretical concepts.

```{r setup}
#| message: FALSE
library(igraph)
```

## Undirected Graph

We'll use an undirected graph with 8 nodes.

```{r}
g <- graph_from_literal(
  1 -- 2, 
  1 -- 3,
  2 -- 4,
  3 -- 5,
  4 -- 5,
  5 -- 6,
  6 -- 7,
  6 -- 8,
  7 -- 8
)

plot(g, vertex.label.cex = 1.2, vertex.size = 20)
```

------------------------------------------------------------------------

###  Degree & Degree Distribution

```{r}
deg <- degree(g)
deg

# Degree distribution
dist <- degree_distribution(g)
plot(dist, type = "h", main = "Degree Distribution", xlab = "Degree", ylab = "Frequency")
```

Identify nodes with the highest and lowest degree. How does the distribution look?

------------------------------------------------------------------------


### Graph Diameter

```{r}
diameter(g)
```

What is the longest shortest path in the graph?

------------------------------------------------------------------------

### Shortest Paths

Find shortest paths from node 1 to all others:

```{r}
sp <- distances(g, v = 1)
sp
```

 Do you understand the output? What is the shortest path from node 1 to node 6?

------------------------------------------------------------------------


### Adjacency Matrix

```{r}
adj_matrix <- as_adjacency_matrix(g, sparse = FALSE)
adj_matrix
```

Use the adjacency matrix to compute the degree of each node. Compare with `degree(g)`.

------------------------------------------------------------------------


### Cutpoints (Articulation Points)

```{r}
articulation_points <- articulation_points(g)
articulation_points
```

### Bridges (Critical Edges)

```{r}
bridge_edges <- which(is.na(edge_connectivity(g)))
E(g)[bridge_edges]

# Alternatively use:
bridges(g)
```

### Visualize Cutpoints and Bridges

```{r}
V(g)$color <- ifelse(V(g) %in% articulation_points, "red", "skyblue")
E(g)$color <- ifelse(E(g) %in% bridges(g), "red", "black")

plot(g, vertex.size = 20, vertex.label.cex = 1.2)
```

 Which nodes and edges are critical to keeping the graph connected?

------------------------------------------------------------------------

## Directed Graph

Now let’s work with a **directed graph**.

```{r}
g_dir <- graph_from_literal(
  A -+ B, A -+ C,
  B -+ D,
  C -+ D,
  D -+ E,
  E -+ F,
  F -+ C
)

plot(g_dir, vertex.label.cex = 1.2, vertex.color = "lightcoral", edge.arrow.size = 0.5)
```


### In-Degree and Out-Degree

```{r}
degree(g_dir, mode = "in")   # incoming links
degree(g_dir, mode = "out")  # outgoing links
```

### Strongly Connected Components

```{r}
components(g_dir, mode = "strong")
```

### Directed Paths and Diameter

```{r}
diameter(g_dir, directed = TRUE)
```

Explore how cycles and direction affect path lengths.

------------------------------------------------------------------------


## Exercises

- Create some other small unidrected and directed graphs and see how the above measure vary on them

- import the Florentine marriage and business network from the `networkdata` package and compute the appropriate measures from above on it
```{r}
library(networkdata)
data("flo_marriage")
data("flo_business")
```
