```{r setup, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

# Cross-Validation and the Bootstrap

In this lab, we explore the resampling techniques covered in this chapter. Some of the commands in this lab may take a while to run on your computer.

```{r chunk1}
library(ISLR2)
set.seed(1)
attach(Auto)
```

## Leave-One-Out Cross-Validation

The LOOCV estimate can be automatically computed for any generalized linear model using the `glm()` and `cv.glm()` functions. In the lab for Chapter 4, we used the `glm()` function to perform logistic regression by passing in the `family = "binomial"` argument. But if we use `glm()` to fit a model without passing in the `family` argument, then it performs linear regression, just like the `lm()` function.

```{r chunk8}
library(boot)
glm.fit <- glm(mpg ~ horsepower, data = Auto)
cv.err <- cv.glm(Auto, glm.fit)
cv.err$delta
```

The `cv.glm()` function produces a list with several components. The two numbers in the `delta` vector contain the cross-validation results. In this case the numbers are identical (up to two decimal places) and correspond to the LOOCV statistic given in (5.1). Below, we discuss a situation in which the two numbers differ. Our cross-validation estimate for the test error is approximately $24.23$.

We can repeat this procedure for increasingly complex polynomial fits. To automate the process, we use the `for()` function to initiate a *for loop* which iteratively fits polynomial regressions for polynomials of order $i=1$ to $i=10$, computes the associated cross-validation error, and stores it in the $i$th element of the vector `cv.error`. We begin by initializing the vector.

```{r chunk9}
cv.error <- rep(0, 10)
for (i in 1:10) {
  glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.error[i] <- cv.glm(Auto, glm.fit)$delta[1]
}
cv.error
```

## $k$-Fold Cross-Validation

The `cv.glm()` function can also be used to implement $k$-fold CV. Below we use $k=10$, a common choice for $k$, on the `Auto` data set. We once again set a random seed and initialize a vector in which we will store the CV errors corresponding to the polynomial fits of orders one to ten.

```{r chunk10}
set.seed(17)
cv.error.10 <- rep(0, 10)
for (i in 1:10) {
  glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.error.10[i] <- cv.glm(Auto, glm.fit, K = 10)$delta[1]
}
cv.error.10
library(tidyverse)
library(ggplot2)
```

```{r}

cvviz = data.frame(err_LOOCV = cv.error, err_K10 = cv.error.10, poly_degree = seq(1:10))
cvviz = cvviz %>%
  pivot_longer(cols = starts_with("err"),
               names_to = "Model",
               values_to = "Error")

cvviz %>%
  ggplot(aes(x = poly_degree, y = Error, color = Model)) +
  geom_point() +
  geom_line() +
  theme_minimal()
```

Notice that the computation time is shorter than that of LOOCV. (In principle, the computation time for LOOCV for a least squares linear model should be faster than for $k$-fold CV, due to the availability of the formula (5.2) for LOOCV; however, unfortunately the `cv.glm()` function does not make use of this formula.) We still see little evidence that using cubic or higher-order polynomial terms leads to lower test error than simply using a quadratic fit.

## The Bootstrap

We illustrate the use of the bootstrap in the simple example of Section 5.2, as well as on an example involving estimating the accuracy of the linear regression model on the `Auto` data set.

### Estimating the Accuracy of a Statistic of Interest

One of the great advantages of the bootstrap approach is that it can be applied in almost all situations. No complicated mathematical calculations are required. Performing a bootstrap analysis in `R` entails only two steps. First, we must create a function that computes the statistic of interest. Second, we use the `boot()` function, which is part of the `boot` library, to perform the bootstrap by repeatedly sampling observations from the data set with replacement.

The `Portfolio` data set in the `ISLR2` package is simulated data of $100$ pairs of returns, generated in the fashion described in Section 5.2. To illustrate the use of the bootstrap on this data, we must first create a function.

`alpha.fn()`, takes as input the $(X,Y)$ data as well as a vector indicating which observations should be used to estimate $\alpha$. The function then outputs the estimate for $\alpha$ based on the selected observations.

```{r chunk11}
alpha.fn <- function(data, index) {
  X <- data$X[index]
  Y <- data$Y[index]
  (var(Y) - cov(X, Y)) / (var(X) + var(Y) - 2 * cov(X, Y))
}
```

This function *returns*, or outputs, an estimate for $\alpha$ based on applying (5.7) to the observations indexed by the argument `index`. For instance, the following command tells `R` to estimate $\alpha$ using all $100$ observations.

```{r chunk12}
alpha.fn(Portfolio, 1:100)
```

The next command uses the `sample()` function to randomly select $100$ observations from the range $1$ to $100$, with replacement. This is equivalent to constructing a new bootstrap data set and recomputing $\hat{\alpha}$ based on the new data set.

```{r chunk13}
set.seed(7)
alpha.fn(Portfolio, sample(100, 100, replace = T))
```

We can implement a bootstrap analysis by performing this command many times, recording all of the corresponding estimates for $\alpha$, and computing the resulting standard deviation. However, the `boot()` function automates this approach. Below we produce $R=1,000$ bootstrap estimates for $\alpha$.

```{r chunk14}
boot(Portfolio, alpha.fn, R = 1000)
```

The final output shows that using the original data, $\hat{\alpha}=0.5758$, and that the bootstrap estimate for ${\rm SE}(\hat{\alpha})$ is $0.0897$.

### Estimating the Accuracy of a Linear Regression Model

The bootstrap approach can be used to assess the variability of the coefficient estimates and predictions from a statistical learning method. Here we use the bootstrap approach in order to assess the variability of the estimates for $\beta_0$ and $\beta_1$, the intercept and slope terms for the linear regression model that uses `horsepower` to predict `mpg` in the `Auto` data set. We will compare the estimates obtained using the bootstrap to those obtained using the formulas for ${\rm SE}(\hat{\beta}_0)$ and ${\rm SE}(\hat{\beta}_1)$ described in Section 3.1.2.

We first create a simple function, `boot.fn()`, which takes in the `Auto` data set as well as a set of indices for the observations, and returns the intercept and slope estimates for the linear regression model. We then apply this function to the full set of $392$ observations in order to compute the estimates of $\beta_0$ and $\beta_1$ on the entire data set using the usual linear regression coefficient estimate formulas from Chapter 3. Note that we do not need the `{` and `}` at the beginning and end of the function because it is only one line long.

```{r chunk15}
boot.fn <- function(data, index)
  coef(lm(mpg ~ horsepower, data = data, subset = index))
boot.fn(Auto, 1:392)
```

The `boot.fn()` function can also be used in order to create bootstrap estimates for the intercept and slope terms by randomly sampling from among the observations with replacement. Here we give two examples.

```{r chunk16}
set.seed(1)
boot.fn(Auto, sample(392, 392, replace = T))
boot.fn(Auto, sample(392, 392, replace = T))
```

Next, we use the `boot()` function to compute the standard errors of 1,000 bootstrap estimates for the intercept and slope terms.

```{r chunk17}
boot(Auto, boot.fn, 1000)
```

This indicates that the bootstrap estimate for ${\rm SE}(\hat{\beta}_0)$ is $0.84$, and that the bootstrap estimate for ${\rm SE}(\hat{\beta}_1)$ is $0.0073$. As discussed in Section 3.1.2, standard formulas can be used to compute the standard errors for the regression coefficients in a linear model. These can be obtained using the `summary()` function.

```{r chunk18}
summary(lm(mpg ~ horsepower, data = Auto))$coef
```

The standard error estimates for $\hat{\beta}_0$ and $\hat{\beta}_1$ obtained using the formulas from Section 3.1.2 are $0.717$ for the intercept and $0.0064$ for the slope. Interestingly, these are somewhat different from the estimates obtained using the bootstrap. Does this indicate a problem with the bootstrap? In fact, it suggests the opposite. Recall that the standard formulas given in Equation 3.8 on page 66 rely on certain assumptions. For example, they depend on the unknown parameter $\sigma^2$, the noise variance. We then estimate $\sigma^2$ using the RSS. Now although the formulas for the standard errors do not rely on the linear model being correct, the estimate for $\sigma^2$ does. We see in Figure 3.8 on page 92 that there is a non-linear relationship in the data, and so the residuals from a linear fit will be inflated, and so will $\hat{\sigma}^2$. Secondly, the standard formulas assume (somewhat unrealistically) that the $x_i$ are fixed, and all the variability comes from the variation in the errors $\epsilon_i$. The bootstrap approach does not rely on any of these assumptions, and so it is likely giving a more accurate estimate of the standard errors of $\hat{\beta}_0$ and $\hat{\beta}_1$ than is the `summary()` function.

Below we compute the bootstrap standard error estimates and the standard linear regression estimates that result from fitting the quadratic model to the data. Since this model provides a good fit to the data (Figure 3.8), there is now a better correspondence between the bootstrap estimates and the standard estimates of ${\rm SE}(\hat{\beta}_0)$, ${\rm SE}(\hat{\beta}_1)$ and ${\rm SE}(\hat{\beta}_2)$.

```{r chunk19}
boot.fn <- function(data, index)
  coef(
      lm(mpg ~ horsepower + I(horsepower^2), 
        data = data, subset = index)
    )
set.seed(1)
boot(Auto, boot.fn, 1000)
summary(
    lm(mpg ~ horsepower + I(horsepower^2), data = Auto)
  )$coef
```

## Exercises

Use the `Boston` data set.

**Important:** Any function that you pass to boot needs to use indeces

(a) Based on this data set, provide an estimate for the population mean of `medv`. Call this estimate $\hat{\mu}$

```{r}


```

(b) Provide an estimate of the standard error of $\hat{\mu}$ . Interpret this result.

    Hint: We can compute the standard error of the sample mean by dividing the sample standard deviation by the square root of the number of observations.

```{r}

```

(c) Now estimate the standard error of $\hat{\mu}$ using the bootstrap. How does this compare to your answer from (b)?

```{r}

```

(d) Based on your bootstrap estimate from (c), provide a 95% confidence interval for the mean of `medv`. Compare it to the results obtained using `t.test(Boston$medv)`.

    Hint: You can approximate a 95% confidence interval using the formula

    \[$\hat{\mu}$ - 2SE($\hat{\mu}$), $\hat{\mu}$ + 2SE($\hat{\mu}$)\] .

```{r}

```

(d) Based on this data set, provide an estimate, $\hat{\mu}_{med}$, for the median value of \`medv\` in the population.

```{r}

```

(d) We now would like to estimate the standard error of $\hat{\mu}_{med}$. Unfortunately, there is no simple formula for computing the standard error of the median. Instead, estimate the standard error of the median using the bootstrap. Comment on your findings.

```{r}

```

(d) Based on this data set, provide an estimate for the tenth per centile of medv in Boston census tracts. Call this quantity $\hat{\mu}_{0.1}$. (You can use the quantile() function.)

```{r}
```

(d) Use the bootstrap to estimate the standard error of $\hat{\mu}_{0.1}$. Comment on your findings.

```{r}

```
