---
title: "Model Selection & Regularization"
author: "Termeh Shafie"
format:
  html:
    embed-resources: true
number-sections: true
toc: true         
editor: visual
execute:
  cache:  true
---

# Regularization
Now that we know how to use model validation to check for overfitting, we might want to help *fix* overfitting. We do this with **regularization** which helps take a model thats to the righthand side of our Bias Variance graph and move it towards the center by making our model *simpler*. 

One way we can do that is by *penalizing* the coefficients of our model if they're too large in magnitude (far away from 0 in either direction). 

## LASSO
LASSO penalizes the sum of the *absolute value* of the coefficients by adding a penalty term to the loss function:

$$ \underbrace{\text{RSS}}_{\sum(x_i - \hat{x_i})^2} + \lambda \sum | \beta_i |$$

LASSO has the benefit of tending to drag coefficients that don't "pull their weight" to *exactly* 0, thus removing them from the model.


## Ridge
Ridge penalizes the sum of the *squared* coefficients by adding a penalty term to the loss function:

$$ \underbrace{\text{RSS}}_{\sum(x_i - \hat{x_i})^2} + \lambda \sum  \beta_i^2$$

Unlike LASSO, Ridge tends to drag coefficients that don't "pull their weight" to *near* 0, thus NOT removing them from the model.


## Penalties in General
Both models create a tug-of-war where coefficients ($\beta_i$) need to "pull their weight" by reducing the Sum of Squared Residuals ($\text{RSS}$) in order to be "worth" having a large value in the penalty. 

The hyperparameter $\lambda$ controls how *much* coefficients are penalized. 

### *Question*
If $\lambda = 0$, what happens to our model?


## Alternative Regularization Explanation

You can also think of Ridge and LASSO as having a "budget" for how big the sum of your (squared or absolute valued) coefficients can be.

<img src="https://drive.google.com/uc?export=view&id=1y_GJ-Y7t2sYU_gVSkJjlYWnnoZmHcTQk" width = "600"/>

This graph shows the concept of LASSO and RIDGE in a simple 2-parameter situation. The teal/blue circle (right) and square (left) represent the values for $\beta_1$ and $\beta_2$ that satisfy the "budget" for our coefficients. The dot in the middle of the red rings represents what the coefficients would be if we did not add a penalty at all. As you move from the inner to the outer red rings, the Residual Sums of Squares (RSS) goes up, meaning that our model is worse at accurately predicting data in our sample (remember that we're giving up accuracy in the *current* sample, to hopefully gain accuracy out-of-sample).

You can see that the coefficients chosen by LASSO/Ridge occur at the point of the teal/blue area that meets the red rings. This is because 1) when we penalize our model, it HAS to fit within the contraints we give it (the teal/blue area) but 2) we still want the RSS to be as small as possible (a model that's bad on ALL data sets isn't useful).

## How to choose between LASSO and Ridge?

In general: if you want to do variable selection (i.e. completely remove some variables), choose LASSO. But in the real world, people often use something called Elastic Net, which adds both LASSO (L1) *and* Ridge (L2) penalties to the loss function (see [here](https://scikit-learn.org/stable/modules/generated/sklearn.linear_model.ElasticNetCV.html)).

# Ridge and LASSO in R
We apply both Ridge and LASSO to the Amazon book data set from before.



```{r}
#| label: load-pckgs
#| message: false
# Packages needed
library(tidyverse)
library(glmnet)
library(knitr)
library(kableExtra)
library(broom)
library(patchwork)
library(leaps) # install if not on your machine
library(ISLR2) # data used from book package
```



Note, when using `glmnet`, the alpha argument determines the penalty term:

- If alpha = 0: pure ridge
- If alpha = 1: pure lasso
- If alpha = 0.5: half ridge + half lasso


## LASSO

```{r}
#| label: lasso

# ---- Load & clean data ----
ama <- read_tsv("07-data/amazon-books.txt") |>
  drop_na()

colnames(ama) <- c(
  "Title", "Author", "List.Price", "Amazon.Price", "Hard.Paper",
  "NumPages", "Publisher", "Pub.year", "ISBN.10",
  "Height", "Width", "Thick", "Weight.oz"
)

# ---- Set up X and y ----
predictors <- c("List.Price", "NumPages", "Weight.oz", "Thick", "Height", "Width")

X <- ama[, predictors]
y <- ama$Amazon.Price

# ---- Train/Test Split (80/20) ----
set.seed(12345)

n <- nrow(ama)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

# Model matrices (glmnet will standardize internally)
X_train_mm <- model.matrix(~ . - 1, data = X_train)
X_test_mm  <- model.matrix(~ . - 1, data = X_test)

# ---- Lasso with Cross-Validation (alpha = 1) ----
cv_fit <- cv.glmnet(
  x = X_train_mm,
  y = y_train,
  alpha = 1,          # Lasso
  standardize = TRUE  # default; explicit for clarity
)

# ---- Predict ----
y_pred_train <- as.numeric(predict(cv_fit, newx = X_train_mm, s = "lambda.min"))
y_pred_test  <- as.numeric(predict(cv_fit, newx = X_test_mm,  s = "lambda.min"))

# ---- Metrics ----
mse  <- function(y, yhat) mean((y - yhat)^2)
mae  <- function(y, yhat) mean(abs(y - yhat))
mape <- function(y, yhat) mean(abs((y - yhat) / y)) * 100
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

# ---- Summary ----
resultsL <- data.frame(
  Dataset = c("Train", "Test"),
  MSE  = c(mse(y_train, y_pred_train),
           mse(y_test,  y_pred_test)),
  MAE  = c(mae(y_train, y_pred_train),
           mae(y_test,  y_pred_test)),
  MAPE = c(mape(y_train, y_pred_train),
           mape(y_test,  y_pred_test)),
  R2   = c(r2(y_train, y_pred_train),
           r2(y_test,  y_pred_test))
)

# ---- Pretty Table ----
resultsL |>
  kable(digits = 4, align = "c",
        caption = "LASSO Regression Performance Summary") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


## Ridge

```{r}
#| label: ridge

# ---- Train/Test Split (80/20) ----
set.seed(12345)  # use same seed as LASSO if you want identical split

n <- nrow(ama)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

X_train <- X[train_idx, ]
X_test  <- X[-train_idx, ]
y_train <- y[train_idx]
y_test  <- y[-train_idx]

# Model matrices (glmnet will standardize internally)
X_train_mm <- model.matrix(~ . - 1, data = X_train)
X_test_mm  <- model.matrix(~ . - 1, data = X_test)

# ---- Ridge with Cross-Validation (alpha = 0) ----
cv_ridge <- cv.glmnet(
  x = X_train_mm,
  y = y_train,
  alpha = 0,          # Ridge
  standardize = TRUE  # default; explicit for clarity
)

# ---- Predict ----
y_pred_train <- as.numeric(predict(cv_ridge, newx = X_train_mm, s = "lambda.min"))
y_pred_test  <- as.numeric(predict(cv_ridge, newx = X_test_mm,  s = "lambda.min"))

# ---- Metrics ----
mse  <- function(y, yhat) mean((y - yhat)^2)
mae  <- function(y, yhat) mean(abs(y - yhat))
mape <- function(y, yhat) mean(abs((y - yhat) / y)) * 100
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

resultsR <- data.frame(
  Dataset = c("Train", "Test"),
  MSE  = c(mse(y_train, y_pred_train),
           mse(y_test,  y_pred_test)),
  MAE  = c(mae(y_train, y_pred_train),
           mae(y_test,  y_pred_test)),
  MAPE = c(mape(y_train, y_pred_train),
           mape(y_test,  y_pred_test)),
  R2   = c(r2(y_train, y_pred_train),
           r2(y_test,  y_pred_test))
)

# ---- Pretty Table ----
resultsR |>
  kable(digits = 4, align = "c",
        caption = "Ridge Regression Performance Summary") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


### Question
Interpret the results from LASSO and Rdige Regression.


::: {.callout-tip collapse="true"}
### Answer
- Ridge Regression shows better test performance than LASSO, with a **lower Test MSE** (4.94 vs 5.39) and **higher Test R²** (0.656 vs 0.624).  
- **LASSO** fits the training data slightly better, but this advantage does not translate into improved test accuracy.  
- The similarity in **MAE** and **MAPE** across models indicates both methods capture overall price patterns comparably.  
- The stronger generalization of Ridge suggests that, for this dataset, **Ridge is the preferred regularization method**, producing more stable estimates without overshrinking important predictors. In other words, Ridge generalizes slightly better than LASSO on this dataset. Ridge has lower Test MSE (4.94 vs 5.39) and higher Test R² (0.656 vs 0.624).
:::


## Plotting Regularization Paths

```{r}
#| label: plot-reg
#| fig-height: 10
#| fig-width: 10
# Fit models
ridge_fit <- glmnet(X, y, alpha = 0)
lasso_fit <- glmnet(X, y, alpha = 1)

# Convert glmnet output to long data frame 
extract_paths <- function(fit, model_name) {
  coefs <- as.matrix(fit$beta)   # p x K (rows = features, cols = lambdas)
  lambdas <- fit$lambda          # length K
  
  # transpose so rows = lambdas, cols = features
  df <- as.data.frame(t(coefs))  # now nrow = length(lambda), ncol = p
  
  df$lambda <- lambdas
  df$model <- model_name
  
  df_long <- df %>%
    tidyr::pivot_longer(
      cols = -c(lambda, model),
      names_to = "term",
      values_to = "estimate"
    ) %>%
    dplyr::mutate(log_lambda = log(lambda))
  
  df_long
}

ridge_df <- extract_paths(ridge_fit, "Ridge")
lasso_df <- extract_paths(lasso_fit, "Lasso")

paths <- bind_rows(ridge_df, lasso_df)

# Ridge plot
g_ridge <- paths %>%
  filter(model == "Ridge") %>%
  ggplot(aes(x = log_lambda, y = estimate, color = term)) +
  geom_line(linewidth = 1) +
  labs(title = "Ridge Regularization Paths",
       x = "log(lambda)", y = "Coefficient") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

# Lasso plot
g_lasso <- paths %>%
  filter(model == "Lasso") %>%
  ggplot(aes(x = log_lambda, y = estimate, color = term)) +
  geom_line(linewidth = 1) +
  labs(title = "Lasso Regularization Paths",
       x = "log(lambda)", y = "Coefficient") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")


# now let's put norms as x axes
paths_norm <- paths %>%
  group_by(model, lambda) %>%
  mutate(
    l1_norm = sum(abs(estimate)),            # For Lasso
    l2_norm = sqrt(sum(estimate^2))          # For Ridge
  ) %>%
  ungroup()

g_ridgenorms <- paths_norm %>%
  filter(model == "Ridge") %>%
  ggplot(aes(x = l2_norm, y = estimate, color = term)) +
  geom_line(linewidth = 1) +
  labs(title = "Ridge Regularization Paths",
       x = "L2 Norm of Coefficients", y = "Coefficient") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

g_lassonorms <- paths_norm %>%
  filter(model == "Lasso") %>%
  ggplot(aes(x = l1_norm, y = estimate, color = term)) +
  geom_line(linewidth = 1) +
  labs(title = "Lasso Regularization Paths",
       x = "L1 Norm of Coefficients", y = "Coefficient") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")


# Combine all
(g_ridge + g_lasso) / (g_ridgenorms + g_lassonorms)

```

Note that we can use `glmnet()` to plot these regularizations plots directly, we do this in the next part. However, the x axis will then look slightly different as `glmnet()` uses norms as x axes, making the plot independent of the specific $\lambda$ scale.

Also note that in the above plot, each model is using its own $\lambda$ sequence generated by `glmnet()`, so:

- Ridge might use $\lambda$ from $0.1 \rightarrow 10^4$
- Lasso might use $\lambda$ from $10^{-4} \rightarrow 10^2$

This is because ridge and lasso get different grids, leading to different x-axis ranges. You can of course set the same grid for both which we do in the next part.

# Classwork: Linear Models and Regularization Methods (from ISLR2)

## Subset Selection Methods

### Best Subset Selection
Here we apply the best subset selection approach to the `Hitters` data from the `ISLR2` package. We wish to predict a baseball player's `Salary` on the basis of various statistics associated with performance in the previous year.

First of all, we note that the `Salary` variable is missing for some of the players. The `is.na()` function can be used to identify the missing observations. It returns a vector of the same length as the input vector, with a `TRUE` for any elements that are missing, and a `FALSE` for non-missing elements. The `sum()` function can then be used to count all of the missing elements.

```{r}
#| label: load-data
names(Hitters)
dim(Hitters)
sum(is.na(Hitters$Salary))
```
Hence we see that `Salary` is missing for $59$ players. The `na.omit()` function removes all of the rows that have missing values in any variable.

```{r}
#| label: remove-nas
Hitters <- na.omit(Hitters)
dim(Hitters)
sum(is.na(Hitters))
```
The `regsubsets()` function (part of the `leaps` library) performs best subset selection by identifying the best model that contains a given number of predictors, where *best* is quantified using RSS. The syntax is the same as for `lm()`. The `summary()` command outputs the best set of variables for each model size.

```{r}
#| label: reg-subset
regfit.full <- regsubsets(Salary ~ ., Hitters)
summary(regfit.full)
```

An asterisk indicates that a given variable is included in the corresponding model. For instance, this output indicates that the best two-variable model contains only `Hits` and `CRBI`. By default, `regsubsets()` only reports results up to the best eight-variable model. But the `nvmax` option can be used in order to return as many variables as are desired. Here we fit up to a 19-variable model.

```{r}
#| label: reg-fit
regfit.full <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19)
reg.summary <- summary(regfit.full)
```

The `summary()` function also returns $R^2$, RSS, adjusted $R^2$, $C_p$, and BIC. We can examine these to try to select the *best* overall model.

```{r}
names(reg.summary)
```

For instance, we see that the $R^2$ statistic increases from $32 \%$, when only one variable is included in the model, to almost $55 \%$, when all variables are included. As expected, the $R^2$ statistic increases monotonically as more variables are included.

```{r}
reg.summary$rsq
```

Plotting RSS, adjusted $R^2$, $C_p$, and BIC for all of the models at once will help us decide which model to select.

```{r}
#| label: plot-sumstats
df = data.frame(adjR2 = reg.summary$adjr2,
                rss = reg.summary$rss,
                Cp = reg.summary$cp,
                BIC = reg.summary$bic,
                Variables = seq(1:19))
df_long = df %>%
  pivot_longer(cols = c(adjR2, rss, Cp, BIC),
               names_to = "Metric",
               values_to = "Value")

df_long %>%
  ggplot(aes(x = Variables, y = Value)) +
  geom_line() +
  facet_wrap(~Metric, scales = "free_y") + 
  labs(x = "Number of Variables", y = "Value") +
  theme_minimal()
```

The `which.max()` function can be used to identify the location of the maximum point of a vector. We will now plot a red dot to indicate the model with the largest adjusted $R^2$ statistic.

```{r}
#| label: max-r2
#| message: false
#| warning: false

max_index <- which.max(df$adjR2)

df %>%
  ggplot(aes(x = Variables, y = adjR2)) +
  geom_line() + 
  geom_point(aes(x = max_index, y = adjR2[max_index]),
             color = "red", size = 4, shape = 20) + 
  labs(x = "Number of Variables", y = "Adjusted R²") +
  theme_minimal()
```

In a similar fashion we can plot the $C_p$ and BIC statistics, and indicate the models with the smallest statistic using `which.min()`.

```{r}
#| label: C_p-BIC
#| message: false
#| warning: false

min_cp <- which.min(df$Cp)
min_bic <- which.min(df$BIC)

df %>%
  ggplot(aes(x = Variables, y = Cp)) +
  geom_line() + 
  geom_point(aes(x = min_cp, y = Cp[min_cp]),
             color = "red", size = 4, shape = 20) + 
  labs(x = "Number of Variables", y = "Cp") +
  theme_minimal()

df %>%
  ggplot(aes(x = Variables, y = BIC)) +
  geom_line() + 
  geom_point(aes(x = min_bic, y = BIC[min_bic]),
             color = "red", size = 4, shape = 20) + 
  labs(x = "Number of Variables", y = "BIC") +
  theme_minimal()
```

The `regsubsets()` function has a built-in `plot()` command which can be used to display the selected variables for the best model with a given number of predictors, ranked according to the BIC, $C_p$, adjusted $R^2$, or AIC. To find out more about this function, type `?plot.regsubsets`.

```{r}
#| label: plot-all
#| message: false
#| warning: false
plot(regfit.full, scale = "r2")
plot(regfit.full, scale = "adjr2")
plot(regfit.full, scale = "Cp")
plot(regfit.full, scale = "bic")
```

The top row of each plot contains a black square for each variable selected according to the optimal model associated with that statistic. For instance, we see that several models share a BIC close to $-150$. However, the model with the lowest BIC is the six-variable model that contains only `AtBat`, `Hits`, `Walks`, `CRBI`, `DivisionW`, and `PutOuts`. We can use the `coef()` function to see the coefficient estimates associated with this model.

```{r}
coef(regfit.full, 6)
```

### Forward and Backward Stepwise Selection

We can also use the `regsubsets()` function to perform forward stepwise or backward stepwise selection, using the argument `method = "forward"` or `method = "backward"`.

```{r}
#| label: forward-backward

regfit.fwd <- regsubsets(Salary ~ ., data = Hitters,
                         nvmax = 19, method = "forward")
summary(regfit.fwd)

regfit.bwd <- regsubsets(Salary ~ ., data = Hitters,
                         nvmax = 19, method = "backward")
summary(regfit.bwd)
```

For instance, we see that using forward stepwise selection, the best one-variable model contains only `CRBI`, and the best two-variable model additionally includes `Hits`. For this data, the best one-variable through six-variable models are each identical for best subset and forward selection. However, the best seven-variable models identified by forward stepwise selection, backward stepwise selection, and best subset selection are different.

```{r}
#| label: 7-var-models

coef(regfit.full, 7)
coef(regfit.fwd, 7)
coef(regfit.bwd, 7)
```

### Choosing Among Models Using the Validation-Set Approach and Cross-Validation

We just saw that it is possible to choose among a set of models of different sizes using $C_p$, BIC, and adjusted $R^2$. We will now consider how to do this using the validation set and cross-validation approaches.

In order for these approaches to yield accurate estimates of the test error, we must use *only the training observations* to perform all aspects of model-fitting---including variable selection. Therefore, the determination of which model of a given size is best must be made using *only the training observations*. This point is subtle but important. If the full data set is used to perform the best subset selection step, the validation set errors and cross-validation errors that we obtain will not be accurate estimates of the test error.

```{r}
#| label: train-test-split

set.seed(1)
train <- sample(c(TRUE, FALSE),
                nrow(Hitters),
                replace = TRUE)

test <- (!train)
```

Now, we apply `regsubsets()` to the training set in order to perform best subset selection.

```{r}
#| label: subset-sel

regfit.best <- regsubsets(Salary ~ .,
                          data = Hitters[train, ],
                          nvmax = 19)
```

Notice that we subset the `Hitters` data frame directly in the call in order to access only the training subset of the data, using the expression `Hitters[train, ]`. We now compute the validation set error for the best model of each model size. We first make a model matrix from the test data.

```{r}
#| label: mod-matrix

test.mat <- model.matrix(Salary ~ ., data = Hitters[test, ])
```

The `model.matrix()` function is used in many regression packages for building an "X" matrix from data. Now we run a loop, and for each size `i`, we extract the coefficients from `regfit.best` for the best model of that size, multiply them into the appropriate columns of the test model matrix to form the predictions, and compute the test MSE.

```{r}
#| label: coeffs

val.errors <- rep(NA, 19)
for (i in 1:19) {
 coefi <- coef(regfit.best, id = i)
 pred <- test.mat[, names(coefi)] %*% coefi
 val.errors[i] <- mean((Hitters$Salary[test] - pred)^2)
}
```

We find that the best model is the one that contains seven variables.

```{r}
#| label: 7-var-model-again
val.errors
which.min(val.errors)
coef(regfit.best, 7)
```

This was a little tedious, partly because there is no `predict()` method for `regsubsets()`. Since we will be using this function again, we can capture our steps above and write our own predict method.

```{r}
#| label: predict-func

 predict.regsubsets <- function(object, newdata, id, ...) {
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form, newdata)
  coefi <- coef(object, id = id)
  xvars <- names(coefi)
  mat[, xvars] %*% coefi
 }
```

Our function pretty much mimics what we did above. The only complex part is how we extracted the formula used in the call to `regsubsets()`. We demonstrate how we use this function below, when we do cross-validation.

Finally, we perform best subset selection on the full data set, and select the best seven-variable model. It is important that we make use of the full data set in order to obtain more accurate coefficient estimates.

Note that we perform best subset selection on the full data set and select the best seven-variable model, rather than simply using the variables that were obtained from the training set, because the best seven-variable model on the full data set may differ from the corresponding model on the training set.

```{r}
#| label: best-subset

regfit.best <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19)
coef(regfit.best, 7)
```

In fact, we see that the best seven-variable model on the full data set has a different set of variables than the best seven-variable model on the training set.

We now try to choose among the models of different sizes using cross-validation. This approach is somewhat involved, as we must perform best subset selection *within each of the* $k$ training sets. Despite this, we see that with its clever subsetting syntax, `R` makes this job quite easy. First, we create a vector that allocates each observation to one of $k=10$ folds, and we create a matrix in which we will store the results.

```{r}
#| label: cv10
k <- 10
n <- nrow(Hitters)
set.seed(1)
folds <- sample(rep(1:k, length = n))
cv.errors <- matrix(NA, k, 19,
    dimnames = list(NULL, paste(1:19)))
```

Now we write a for loop that performs cross-validation. In the $j$th fold, the elements of `folds` that equal `j` are in the test set, and the remainder are in the training set. We make our predictions for each model size (using our new `predict()` method), compute the test errors on the appropriate subset, and store them in the appropriate slot in the matrix `cv.errors`. Note that in the following code `R` will automatically use our `predict.regsubsets()` function when we call `predict()` because the `best.fit` object has class `regsubsets`.

```{r}
#| label: cv-loop
for (j in 1:k) {
  best.fit <- regsubsets(Salary ~ .,
       data = Hitters[folds != j, ],
       nvmax = 19)
  for (i in 1:19) {
    pred <- predict(best.fit, Hitters[folds == j, ], id = i)
    cv.errors[j, i] <-
         mean((Hitters$Salary[folds == j] - pred)^2)
   }
 }
```

This has given us a $10 \times 19$ matrix, of which the $(j,i)$th element corresponds to the test MSE for the $j$th cross-validation fold for the best $i$-variable model. We use the `apply()` function to average over the columns of this matrix in order to obtain a vector for which the $i$th element is the cross-validation error for the $i$-variable model.

```{r}
#| label: cv-results
mean.cv.errors <- apply(cv.errors, 2, mean)
mean.cv.errors
par(mfrow = c(1, 1))
plot(mean.cv.errors, type = "b")
```

We see that cross-validation selects a 10-variable model. We now perform best subset selection on the full data set in order to obtain the 10-variable model.

```{r}
#| label: best-choice

reg.best <- regsubsets(Salary ~ ., data = Hitters,
    nvmax = 19)
coef(reg.best, 10)
```

## Ridge Regression and the Lasso

We will use the `glmnet` package in order to perform ridge regression and the lasso. The main function in this package is `glmnet()`, which can be used to fit ridge regression models, lasso models, and more. This function has slightly different syntax from other model-fitting functions that we have encountered thus far in this book. In particular, we must pass in an `x` matrix as well as a `y` vector, and we do not use the `y ~ x` syntax. We will now perform ridge regression and the lasso in order to predict `Salary` on the `Hitters` data. Before proceeding ensure that the missing values have been removed from the data, as described in Section 6.5.1.

```{r}
#| label: vars-definitions
x <- model.matrix(Salary ~ ., Hitters)[, -1]
y <- Hitters$Salary
```

The `model.matrix()` function is particularly useful for creating `x`; not only does it produce a matrix corresponding to the $19$ predictors but it also automatically transforms any qualitative variables into dummy variables. The latter property is important because `glmnet()` can only take numerical, quantitative inputs.

### Ridge Regression

The `glmnet()` function has an `alpha` argument that determines what type of model is fit. If `alpha=0` then a ridge regression model is fit, and if `alpha=1` then a lasso model is fit. We first fit a ridge regression model.

```{r}
#| label: ridge-hitters
grid <- 10^seq(10, -2, length = 100)
ridge.mod <- glmnet(x, y, alpha = 0, lambda = grid)
```

By default the `glmnet()` function performs ridge regression for an automatically selected range of $\lambda$ values. However, here we have chosen to implement the function over a grid of values ranging from $\lambda=10^{10}$ to $\lambda=10^{-2}$, essentially covering the full range of scenarios from the null model containing only the intercept, to the least squares fit. As we will see, we can also compute model fits for a particular value of $\lambda$ that is not one of the original `grid` values. Note that by default, the `glmnet()` function standardizes the variables so that they are on the same scale. To turn off this default setting, use the argument `standardize = FALSE`.

Associated with each value of $\lambda$ is a vector of ridge regression coefficients, stored in a matrix that can be accessed by `coef()`. In this case, it is a $20 \times 100$ matrix, with $20$ rows (one for each predictor, plus an intercept) and $100$ columns (one for each value of $\lambda$).

```{r}
dim(coef(ridge.mod))
```

We expect the coefficient estimates to be much smaller, in terms of $\ell_2$ norm, when a large value of $\lambda$ is used, as compared to when a small value of $\lambda$ is used. These are the coefficients when $\lambda=11{,}498$, along with their $\ell_2$ norm:

```{r}
ridge.mod$lambda[50]
coef(ridge.mod)[, 50]
sqrt(sum(coef(ridge.mod)[-1, 50]^2))
```

In contrast, here are the coefficients when $\lambda=705$, along with their $\ell_2$ norm. Note the much larger $\ell_2$ norm of the coefficients associated with this smaller value of $\lambda$.

```{r}
ridge.mod$lambda[60]
coef(ridge.mod)[, 60]
sqrt(sum(coef(ridge.mod)[-1, 60]^2))
```

We can use the `predict()` function for a number of purposes. For instance, we can obtain the ridge regression coefficients for a new value of $\lambda$, say $50$:

```{r}
#| label: ridge-predict
predict(ridge.mod, s = 50, type = "coefficients")[1:20, ]
```

We now split the samples into a training set and a test set in order to estimate the test error of ridge regression and the lasso. There are two common ways to randomly split a data set. The first is to produce a random vector of `TRUE`, `FALSE` elements and select the observations corresponding to `TRUE` for the training data. The second is to randomly choose a subset of numbers between $1$ and $n$; these can then be used as the indices for the training observations. The two approaches work equally well. We used the former method in Section 6.5.1. Here we demonstrate the latter approach.

We first set a random seed so that the results obtained will be reproducible.

```{r}
#| label: split-data
set.seed(1)
train <- sample(1:nrow(x), nrow(x) / 2)
test <- (-train)
y.test <- y[test]
```

Next we fit a ridge regression model on the training set, and evaluate its MSE on the test set, using $\lambda=4$. Note the use of the `predict()` function again. This time we get predictions for a test set, by replacing `type="coefficients"` with the `newx` argument.

```{r}
#| label: ridge-train-test
ridge.mod <- glmnet(x[train, ], y[train], alpha = 0,
    lambda = grid, thresh = 1e-12)

ridge.pred <- predict(ridge.mod, s = 4, newx = x[test, ])
mean((ridge.pred - y.test)^2)
```

The test MSE is $142{,}199$. Note that if we had instead simply fit a model with just an intercept, we would have predicted each test observation using the mean of the training observations. In that case, we could compute the test set MSE like this:

```{r}
mean((mean(y[train]) - y.test)^2)
```

We could also get the same result by fitting a ridge regression model with a *very* large value of $\lambda$. Note that `1e10` means $10^{10}$.

```{r}
#| label: large-lambda
ridge.pred <- predict(ridge.mod, s = 1e10, newx = x[test, ])
mean((ridge.pred - y.test)^2)
```

So fitting a ridge regression model with $\lambda=4$ leads to a much lower test MSE than fitting a model with just an intercept. We now check whether there is any benefit to performing ridge regression with $\lambda=4$ instead of just performing least squares regression. Recall that least squares is simply ridge regression with $\lambda=0$. (In order for `glmnet()` to yield the exact least squares coefficients when $\lambda=0$, we use the argument `exact = T` when calling the `predict()` function. Otherwise, the `predict()` function will interpolate over the grid of $\lambda$ values used in fitting the `glmnet()` model, yielding approximate results. When we use `exact = T`, there remains a slight discrepancy in the third decimal place between the output of `glmnet()` when $\lambda = 0$ and the output of `lm()`; this is due to numerical approximation on the part of `glmnet()`.)

```{r}
#| label: ridge-preds
ridge.pred <- predict(ridge.mod, s = 0, newx = x[test, ],
    exact = T, x = x[train, ], y = y[train])
mean((ridge.pred - y.test)^2)
lm(y ~ x, subset = train)
predict(ridge.mod, s = 0, exact = T, type = "coefficients",
    x = x[train, ], y = y[train])[1:20, ]
```

In general, if we want to fit a (unpenalized) least squares model, then we should use the `lm()` function, since that function provides more useful outputs, such as standard errors and p-values for the coefficients.

In general, instead of arbitrarily choosing $\lambda=4$, it would be better to use cross-validation to choose the tuning parameter $\lambda$. We can do this using the built-in cross-validation function, `cv.glmnet()`. By default, the function performs ten-fold cross-validation, though this can be changed using the argument `nfolds`. Note that we set a random seed first so our results will be reproducible, since the choice of the cross-validation folds is random.

```{r}
#| label: cv-ridge

set.seed(1)
cv.out <- cv.glmnet(x[train, ], y[train], alpha = 0)
plot(cv.out)
bestlam <- cv.out$lambda.min
bestlam
```

Therefore, we see that the value of $\lambda$ that results in the smallest cross-validation error is $326$. What is the test MSE associated with this value of $\lambda$?

```{r}
#| label: cv-ridge-preds

ridge.pred <- predict(ridge.mod, s = bestlam,
    newx = x[test, ])
mean((ridge.pred - y.test)^2)
```

This represents a further improvement over the test MSE that we got using $\lambda=4$. Finally, we refit our ridge regression model on the full data set, using the value of $\lambda$ chosen by cross-validation, and examine the coefficient estimates.

```{r}
#| label: coeffs-final
out <- glmnet(x, y, alpha = 0)
predict(out, type = "coefficients", s = bestlam)[1:20, ]
```

As expected, none of the coefficients are zero---ridge regression does not perform variable selection!

### The Lasso

We saw that ridge regression with a wise choice of $\lambda$ can outperform least squares as well as the null model on the `Hitters` data set. We now ask whether the lasso can yield either a more accurate or a more interpretable model than ridge regression. In order to fit a lasso model, we once again use the `glmnet()` function; however, this time we use the argument `alpha=1`. Other than that change, we proceed just as we did in fitting a ridge model.

```{r}
#| message: false
#| warning: false
#| label: lasso-hitters
lasso.mod <- glmnet(x[train, ], y[train], alpha = 1,
    lambda = grid)
plot(lasso.mod)
```

We can see from the coefficient plot that depending on the choice of tuning parameter, some of the coefficients will be exactly equal to zero. We now perform cross-validation and compute the associated test error.

```{r}
#| label: cv-lasso
#| message: false
#| warning: false
set.seed(1)
cv.out <- cv.glmnet(x[train, ], y[train], alpha = 1)
plot(cv.out)
bestlam <- cv.out$lambda.min
lasso.pred <- predict(lasso.mod, s = bestlam,
    newx = x[test, ])
mean((lasso.pred - y.test)^2)
```

This is substantially lower than the test set MSE of the null model and of least squares, and very similar to the test MSE of ridge regression with $\lambda$ chosen by cross-validation.

However, the lasso has a substantial advantage over ridge regression in that the resulting coefficient estimates are sparse. Here we see that 8 of the 19 coefficient estimates are exactly zero. So the lasso model with $\lambda$ chosen by cross-validation contains only eleven variables.

```{r}
#| label: lasso-vars-selection
out <- glmnet(x, y, alpha = 1, lambda = grid)
lasso.coef <- predict(out, type = "coefficients",
    s = bestlam)[1:20, ]
lasso.coef
lasso.coef[lasso.coef != 0]
```

