---
author:  Termeh Shafie
title: "Social Network Analysis"
subtitle: "Worksheet 1: Introduction"
lightbox: auto
toc: true
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 10
---

## R packages for network analysis 

Throughout the course we will use a variety of different packages of doing network analysis, modeling and visualization. Make sure to install them all and have them ready to load when needed:

```{r install_packages}
#| eval: FALSE
install.packages("igraph")   
install.packages("statnet")  #installs ergm, network, and sna
install.packages("snahelper")
install.packages("netUtils")
install.packages("ggraph")
install.packages("backbone")
install.packages("netrankr")
install.packages("signnet")
install.packages("intergraph")
install.packages("graphlayouts")
install.packages("visNetwork")
install.packages("patchwork")
install.packages("edgebundle")
install.packages("ggplot2")
install.packages("gganimate")
install.packages("ggforce")
install.packages("rsiena")

install.packages("remotes")
remotes::install_github("schochastics/networkdata")
```

### Load Packages
We will start with loading the following packages:
```{r load_packages}
#| code-fold: false
#| warning: false
library(igraph)
library(networkdata)
library(netUtils)
```


### Handling package conflicts

**never load sna and igraph at the same time**
```{r load_sna}
#| code-fold: false
#| warning: false
library(sna)
```

```{r error}
#| error: true
#| code-fold: false
data("flo_marriage")
degree(flo_marriage)
```

If for any reason you have to load both, you can circumvent the error by explicitly stating package first 
```{r solution}
igraph::degree(flo_marriage)
```

The package `intergraph` can be used to transform an `igraph` object into a network object and vice versa.

```{r transform_i}
#| code-fold: false
#install.packages("intergraph")
library(intergraph)
asNetwork(flo_marriage)
degree(asNetwork(flo_marriage))
```


You can unload a package without restarting R/RStudio.
```{r detach_sna}
detach("package:sna", unload = TRUE)
```
